---
title: "barbell lift performance analysis & prediction based on accelerometer data"
author: "Johannes"
date: "27 7 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis shall develop a prediction algorithm that predicts the performance of barbell lifts. 
it follows the following steps:
1. getting and cleaning data 
2. exploratory data analysis
3. predictor- and model selection
4. model optimization and prediction


## 1. getting and cleaning data
```{r get data}
wbdat <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
wbval <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```
The data contains data from 3 accelerometers:
1. belt
2. arm
3. forearm
for severeal persons "User_name"
The class variable indicates the performance of the barbell lift. 

We can clean the data by taking converting the class variable to a factor and removing the NA data which is in the columns of the variance.  
```{r clean data}
wbdat$classe <- as.factor(wbdat$classe)
wbdat$user_name <- as.factor(wbdat$user_name)
wbval$user_name <- as.factor(wbval$user_name)

ixAcc <- grep("acc", colnames(wbdat))
wbdatAc <- cbind(wbdat[,ixAcc], user_name = wbdat$user_name, classe = wbdat$classe)
wbvalAc <- cbind(wbval[,ixAcc], user_name = wbval$user_name, problem_id = wbval$problem_id)
ixvar <- grep("var", colnames(wbdatAc))
wbdatAc <- wbdatAc[,-ixvar]
wbvalAc <- wbvalAc[,-ixvar]
summary(wbdatAc)
```
## 2. exploratory data analysis
Let's have a look on the variables and their relevance using the 

```{r echo=TRUE}
library(caret)
library(ggplot2)
nearZeroVar(wbdatAc, saveMetrics = TRUE)
```
Since the unique percentage of the total_accel... values is below the value of the single accelerometer data we start analyzing based on those predictors.


```{r echo=FALSE}
par(mfrow=c(2,3))
qplot(accel_dumbbell_x, total_accel_arm, data = wbdatAc, colour=classe)
qplot(accel_dumbbell_x, total_accel_forearm, data = wbdatAc, colour=classe)
qplot(accel_dumbbell_x, total_accel_belt, data = wbdatAc, colour=classe)
qplot(total_accel_arm, total_accel_forearm, data = wbdatAc, colour=classe)
qplot(total_accel_arm, total_accel_belt, data = wbdatAc, colour=classe)
qplot(total_accel_forearm, total_accel_belt, data = wbdatAc, colour=classe)
```
As result we can observe clear patterns that we try to exploit in the next steps.

## 3. predictor- and model selection
We will split the data into a training and a test set

```{r}
inTrain <- createDataPartition(y=wbdatAc$classe, p = 0.7, list = FALSE)
wbtrain <- wbdatAc[inTrain, ]
wbtest <- wbdatAc[-inTrain, ]
```

Now we start applying the models comparing the accuracy of each model
1. Linear discriminant analysis
```{r echo=TRUE}

control <- trainControl(method="cv", number=3)
m0 <- train(classe~., data = wbtrain, method = "lda", trainControl = control) 
pred1 <- predict(m0, wbtest)
confusionMatrix(pred1, wbtest$classe)$overall[1]
```
2. K nearest neighbour
```{r echo=TRUE}
m01 <- train(classe~., data=wbtrain, method = "knn")
pred01 <- predict(m01, wbtest)
confusionMatrix(pred01, wbtest$classe)$overall[1]
```
3. Random Forest
```{r echo = TRUE}
library(randomForest)
m3 <- randomForest(classe~., data=wbtrain)
pred3 <- predict(m3, wbtest)
confusionMatrix(pred3, wbtest$classe)$overall[1]
```
The complete confusion matrix is:
```{r}
confusionMatrix(pred3, wbtest$classe)
```

We will select the random forest model for the further predictions.

## 4. Apply to validation/"Quizz" set
```{r}
predval <- predict(m3, wbvalAc)
data.frame(wbvalAc$problem_id, predval)
```